### 2020-09-08 text/article_vedrovice.Rmd
minor editions in spatial analysis

The Cemetery of Vedrovice is one of the earliest and most intensively studied
*Linearbandkeramik* (LBK) cemeteries.
Recent contributions use the evidence from the cemetery in a broader discussion
of post-marital residence patterns [@hrncir2020] and
sexual inequalities and gender in the Neolithic [@masclanslatorre2020; @bickle2020].
<!-- Here we focus on detailed analysis of the distribution of the grave goods and
spatial analysis of the Vedrovice cemetery in respect to gender relations and
social structure. -->
Here we explore the relations between burials and bodies in terms of biologicall
sex, age, localness, exceptionality of grave goods and position within the cemetery.

<!-- view of gender in the Neolithic, archaeological context -->
@robb2017 [128] noted lack of gender archaeology debate for the
European Neolithic in comparison with other periods.
This is due to the notion of gender in the Neolihic that is presented as being
of a very different nature than gender in the later prehistory.
Authors show that gender in the Neolithic is not clearly expressed and
biological differences between sexes are not transformed into binary cultural
categories (ibid. 140-142).
Considering the Neolithic burial evidence across Europe, authors find that it
'shows very little gender distinction; when it does, it is rarely a clear-cut
binary categorization but a matter of degree and overlap, sometimes polythetically
combined with factors such as age and localness' (ibid. 138).
In response, @bickle2020 [14] argues that 'biological sex may not have defined gender
during the LBK' although the 'sexed bodies did matter'.

<!-- gender relations in LBK -->
Although gender relations in the LBK might not be

@@

### Questions

* Is there evidence that certain artefacts covary,
    either positively or negatively?
    Do certain artefacts tend to occur together, or are certain combinations
    absent or rare?
* How can the presence and absence data best display which burials
    contain similar artefacts,
    and which artefacts have similar distributions over burials?
* *How can the data be analysed so as to show most clearly the relationships*
    *between the different* **burials** in terms of their **contents**?
* *How can the data be analysed so as to show most clearly the relationship*
    *between the different* **artefacts** in terms of their
    **distributions in graves**?
* How are grave good assemblages related to age and sex?
(... variations in burial rite coinciding with age or sex of the deceased ...)

@@

<!-- sample (burials, artefacts) -->
Originally, the burial ground consisted of at least 108 burials.
Some of the burials were disturbed by later settlement and burial pits in the
area of the cemetery, amateur excavations during the 19th century or later
agricultural activities.
These burials are excluded from the analysis due to missing and/or
incomplete data and only a sample of 78 well-preserved burials is analyzed here.
Most of the data was obtained from the original publication [@podborsky2002],
anthropological determination of sex and age were corrected based on
more recent analysis by @dockalova2008. Carbon, nitrogen and strontium isotope
ratios were collected in @whittle2013 and @richards2008.
Further details on use and function of the artefacts were gathered from
@masclanslatorre2020.

@@

<!-- To characterize the pattern based on density, the quadrat test is employed.
It divides the area of interest into several cells of fixed sizes
(quadrats) and compares the count of events per quadrat with the count
expected under the null hypothesis that the underlying point process
is completely spatially random (approximated by a Poisson distribution). -->
Functions $G$, $F$, Ripley's $K$ [@ripley1976; @sayer2013] and $T$ [@schladitz2000] are
based on the nearest neighbour distances and are commonly used
to characterize point patterns in archaeology [@nakoinz2016, 135-144].
The $G$ function is a metric of how close events are and summarises
cummulative frequency distribution of the nearest neighbour distances.
Because of its dependence on the intensity of events in the study area, it is
subject to changes in the size of the study area (window).
The $F$ function, an empty-space function, is based on simulating a point pattern
and calculating of distances between random locations and the observed locations.
It is a measure of how far are observed events from any arbitrary location [@osullivan2010, 132-135].
Ripley's $K$ function is in comparison with the $G$ function resilient to
changes in the window size and works with all the proximal events up to a
certain threshold, not only the nearest neighbours [@nakoinz2016, 138].
It is widely used in archaeology cf. [@negre2018].
Last but not least, the $T$ function is a third-order parallel to the $K$ function [@schladitz2000].
For formulas and details on individual functions see the supplement.

Given statistics for independent random process are estimated using Monte Carlo
simulation with 999 iterations.
Although much smaller number of permutations should be sufficient
[@baddeley2016, 384] same as @sayer2013 [78] we chose to use 999 permutations.
Ten most extreme realizations of the simulation are discarded as outliers to
get a significance level $\alpha = 0.01$ [@baddeley2016 410].

To explore whether neighboring individuals are more often of the same sex we use
multiple methods to establish neighborhood relations.
The first two methods are based on constructing neighborhood graphs.
We use Delaunay triangulation and its subgraph, a Gabriel graph [@gabriel1969]
with burial locations as nodes (Fig. \@ref(fig:plan-graphs)).
The Delaunay triangulation can be considered a maximum number of neighbors case
scenario while the Gabriel graph the opposite.
For each node, mean number of neighbors of a given sex is determined and recorded.
To check how common the observed value is, a permutation test
where sex of burials is randomly assigned while preserving the overall
proportion of given categories and the neighborhood structure determined by the
graphs is performed.
The observed values of neighbors sex categories are compared against distributions
derived from the simulated data.

Another approach of establishing neigborhood is by constructing buffer zones
around each of the burials with 6 meters radius.
The size of the radius is based on the results of nearest neighbors analysis, where
at the given radius there is strong evidence of a clustering process (Fig. \@ref(fig:spat-funs)).
The randomness is again tested using a permutation test similar to the one
described for the neighborhood derived from the graphs.

@@

<!-- software packages -->
Data is analyzed in an R environment for statistical computing and graphics
[@rcore] using functions of many different packages, for the complete list
see the supplementary information.

>>>>

The Cemetery of Vedrovice is one of the earliest and most intensively studied
*Linearbandkeramik* (LBK) cemeteries.
Recent contributions use the evidence from the cemetery in a broader discussion
of post-marital residence patterns [@hrncir2020] and
sexual inequalities and gender in the Neolithic [@masclanslatorre2020; @bickle2020].
<!-- Here we focus on detailed analysis of the distribution of the grave goods and
spatial analysis of the Vedrovice cemetery in respect to gender relations and
social structure. -->

<!-- view of gender in the Neolithic, archaeological context -->
@robb2017 [128] noted lack of gender archaeology debate for the
European Neolithic in comparison with other periods.
Authors present gender in the Neolihic as being
of a very different nature than gender in the later prehistory.
It is shown that gender in the Neolithic is not clearly expressed and
biological differences between sexes are not transformed into binary cultural
categories (ibid. 140-142).
Considering the Neolithic burial evidence across Europe, authors find that it
'shows very little gender distinction; when it does, it is rarely a clear-cut
binary categorization but a matter of degree and overlap, sometimes polythetically
combined with factors such as age and localness' (ibid. 138).
In response, @bickle2020 [14] argues that 'biological sex may not have defined gender
during the LBK' although the 'sexed bodies did matter'.

Here we explore the relations between burials and bodies in terms of biological
sex, age, local origin, exceptionality of grave goods and location within the
cemetery.
We focus predominantly on investigating non-random patterns in artefact
co-occurences and spatial arragement of the cemetery in relation to the
proposed *levels* of neolithic identity, i.e. sex, age and localness.
Our question is thus how are grave good assemblages related to aforementioned
categories and whether there is a consistent relationship between
these categories forming the *identity*, grave goods and spatial arrangement
of the cemetery.

<!-- gender relations in LBK -->
<!-- Although gender relations in the LBK might not be -->

@@

<!-- ### Questions
* Is there evidence that certain artefacts covary,
    either positively or negatively?
    Do certain artefacts tend to occur together, or are certain combinations
    absent or rare?
* How can the presence and absence data best display which burials
    contain similar artefacts,
    and which artefacts have similar distributions over burials?
* *How can the data be analysed so as to show most clearly the relationships*
    *between the different* **burials** in terms of their **contents**?
* *How can the data be analysed so as to show most clearly the relationship*
    *between the different* **artefacts** in terms of their
    **distributions in graves**?
* How are grave good assemblages related to age and sex?
(... variations in burial rite coinciding with age or sex of the deceased ...) -->

@@

<!-- sample (burials, artefacts) -->
Originally, the burial ground consisted of at least 108 burials.
Some of the burials were disturbed by later settlement and burial pits in the
area of the cemetery, amateur excavations during the 19th century or later
agricultural activities.
These burials are excluded from the analysis due to missing and/or
incomplete data and only a sample of 78 well-preserved burials is analyzed here.
Most of the data was obtained from the original publication [@podborsky2002],
anthropological determination of sex and age were corrected based on
more recent analysis by @dockalova2008. Carbon, nitrogen and strontium isotope
ratios were collected in @whittle2013 and @richards2008.
Further details on use and function of the artefacts were gathered from
@masclanslatorre2020.
Information on localness of different bodies was derived from @hrncir2020.

@@

<!-- To characterize the pattern based on density, the quadrat test is employed.
It divides the area of interest into several cells of fixed sizes
(quadrats) and compares the count of events per quadrat with the count
expected under the null hypothesis that the underlying point process
is completely spatially random (approximated by a Poisson distribution). -->
Functions *G*, *F*, Ripley's *K* [@ripley1976; @sayer2013] and *T* [@schladitz2000] are
based on the nearest neighbour distances and are commonly used
to characterize point patterns in archaeology [@nakoinz2016, 135-144].
The *G* function is a metric of how close events are and summarises
cummulative frequency distribution of the nearest neighbour distances.
Because of its dependence on the intensity of events in the study area, it is
subject to changes in the size of the study area (window).
The *F* function, an empty-space function, is based on simulating a point pattern
and calculating of distances between random locations and the observed locations.
It is a measure of how far observed events are from any arbitrary location [@osullivan2010, 132-135].
Ripley's *K* function is in comparison with the *G* function resilient to
changes in the window size and works with all the proximal events up to a
certain threshold, not only the nearest neighbours [@nakoinz2016, 138].
It is widely used in archaeology cf. [@negre2018].
Last but not least, the *T* function is a third-order parallel to the *K* function [@schladitz2000].
For formulas and details on individual functions see the supplement.

Given statistics for independent random process are estimated using Monte Carlo
simulation with 999 iterations.
Although much smaller number of permutations should be sufficient
[@baddeley2016, 384] same as @sayer2013 [78] we chose to use 999 permutations.
Ten most extreme realizations of the simulation are discarded as outliers so
the significance level $\alpha = 0.01$ [@baddeley2016 410].

To explore whether neighboring individuals are more often of the same sex we use
multiple methods to establish neighborhood relations.
The first two methods are based on constructing neighborhood graphs.
We use Delaunay triangulation and its subgraph, a Gabriel graph [@gabriel1969]
with burial locations as nodes (Fig. \@ref(fig:plan-graphs)).
The Delaunay triangulation can be considered a maximum number of neighbors case
scenario while the Gabriel graph the opposite.
For each node, mean number of neighbors of a given category is determined and recorded.
To check how common the observed value is, a permutation test is performed
where categories are randomly assigned while preserving the overall
proportion of given categories and the neighborhood structure determined by the
graph.
The observed values of neighbors categories are compared against distributions
derived from 999 permutations of the simulated data.

Another approach considered for establishing neigborhood relations is by
constructing buffer zones around each of the burials.
The question arises how large the buffer zones should be.
To solve this problem we utilise percolation analysis [@maddison2020].
Percolation analysis is a method used to identify clusters of points based
on Euclidean distance.

@maddison2020

The size of the radius is based on the results of nearest neighbors analysis, where
at the given radius there is strong evidence of a clustering process (Fig. \@ref(fig:spat-funs)).
The randomness is again tested using a permutation test similar to the one
described for the neighborhood derived from the graphs.

@@

<!-- software packages -->
Data is analyzed in an R environment for statistical computing and graphics
[@rcore] using functions of several different packages, for the complete list
see the supplementary information.
